Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-02-06 WSSSPE2_report/WSSSPE2_review.tex
maybe now ready for others to work

The Second Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} will be held on Sunday, 16 November 2014, in conjunction with the 2014 International Conference for High Performance Computing, Networking, Storage and Analysis (SC14)\footnote{\url{http://sc14.supercomputing.org}}.
WSSSPE2 follows a general initial workshop, WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1}, that was held in 2013 in conjunction with the SC13 conference, and a focused workshop, WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, that was held in 2014 in conjunction with the SciPy conference.

@@

\section{Submissions}

@@

\section{Peer-Review and Peer-Grouping}

@@

The organizers then chose the soft with five groups as the one that felt most meaningful, then decided on names for the five groups, namely:
\begin{itemize}
\item Exploring Sustainability
\item Software Development Experiences
\item Credit \& Incentives
\item Reproducibility \& Reuse \& Sharing
\item Code Testing \& Code Review
\end{itemize}

Finally, since some of the papers were not represented by cards in the process, they were not placed in groups, so the authors of these papers were asked which group seemed the best for their papers, and those papers were then placed in those groups, as listed in the next section of this report.

\section{Results}

The contributed papers that will be discussed at the workshop follow, listed by groups (determined as described in the previous section.)

\begin{itemize}
\item Exploring Sustainability
\begin{itemize}
\item Mario {Rosado de Souza} and Robert Haines and Caroline Jay. Defining Sustainability through Developers' Eyes: Recommendations from an Interview Study~\cite{wssspe2_rosada_de_souza}
\item Robert Downs, W. Christopher Lenhardt, Erin Robinson, Ethan Davis, Nicholas Weber. Community Recommendations for Sustainable Scientific Software~\cite{wssspe2_downs}
\item Abani Patra, Matthew Jones, Steven Gallo, Kyle Marcus and Tevfik Kosar. Role of Online Platforms, Communications and Workflows in Developing Sustainable Software for Science Communities~\cite{wssspe2_patra}
\item Marlon Pierce, Suresh Marru and Chris Mattmann.{WSSSPE2}: Patching It Up, Pulling It Forward~\cite{wssspe2_pierce}
\item Justin Shi. Seeking the Principles of Sustainable Software Engineering~\cite{wssspe2_shi}
\item Colin C. Venters, Michael K. Griffiths, Violeta Holmes, Rupert R. Ward and David J. Cooke. The Nebuchadnezzar Effect: Dreaming of Sustainable Software through Sustainable Software Architectures~\cite{wssspe2_venters}
\end{itemize}

\item Software Development Experiences
\begin{itemize}
\item Jordan Adams, Sai Nudurupati, Nicole Gasparini, Daniel Hobley, Eric Hutton, Gregory Tucker and Erkan Istanbulluoglu. Landlab: Sustainable Software Development in Practice ~\cite{wssspe2_adams}
\item Alice Allen and Judy Schmidt. Looking before leaping: Creating a software registry~\cite{wssspe2_allen}
\item Carl Boettiger, Ted Hart, Scott Chamberlain and Karthik Ram. Building software, building community: lessons from the {ROpenSci} project~\cite{wssspe2_boettiger}
\item Michael R. Crusoe and C.Titus Brown. Channeling community contributions to scientific software: a hackathon experience~\cite{wssspe2_crusoe}
\item Yolanda Gil,  Eunyoung Moon and James Howison. No Science Software is an Island: Collaborative Software Development Needs in Geosciences~\cite{wssspe2_gil}
\item Ted Habermann, Andrew Collette, Steve Vincena, Werner Benger, Jay Jay Billings, Matt Gerring, Konrad Hinsen, Pierre de Buyl, Mark K\"{o}nnecke, Filipe Rnc Maia and Suren Byna. The Hierarchical Data Format ({HDF}): A Foundation for Sustainable Data and Software~\cite{wssspe2_habermann}
\item Marcus Hanwell, Patrick O'Leary and Bob O'Bara. Sustainable Software Ecosystems: Software Engineers, Domain Scientists, and Engineers Collaborating for Science~\cite{wssspe2_hanwell}
\item Eric Hutton, Mark Piper, Irina Overeem, Albert Kettner and James Syvitski. Building Sustainable Software - The {CSDMS} Approach~\cite{wssspe2_hutton}
\item W. Christopher Lenhardt, Stanley Ahalt, Matt Jones, J. Aukema, S. Hampton, S. R. Hespanh, R. Idaszak and M. Schildhauer. {ISEES-WSSI} Lessons for Sustainable Science Software from an Early Career Training Institute on Open Science Synthesis~\cite{wssspe2_lenhardt}
\item Jory Schossau and Greg Wilson. Which Sustainable Software Practices Do Scientists Find Most Useful?~\cite{wssspe2_schossau}
\item James S. Spencer, Nicholas S. Blunt, William A. Vigor, Fionn D. Malone, W. M. C. Foulkes, James J. Shepherd and Alex J. W. Thom. The {H}ighly {A}ccurate {N-DE}terminant ({HANDE}) quantum {Monte Carlo} project: Open-source stochastic diagonalisation for quantum chemistry~\cite{wssspe2_spencer}
\end{itemize}

\item Credit \& Incentives
\begin{itemize}
\item James Howison. Retract bit-rotten publications: Aligning incentives for sustaining scientific software~\cite{wssspe2_howison}
\item Daniel Katz and Arfon Smith. Implementing Transitive Credit with {JSON-LD}~\cite{wssspe2_katz}
\item Ian Kelley. Publish or perish: the credit deficit to making software and generating data~\cite{wssspe2_kelley}
\end{itemize}

\item Reproducibility \& Reuse \& Sharing
\begin{itemize}
\item Jakob Blomer, Dario Berzano, Predrag Buncic, Ioannis Charalampidis, Gerardo Ganis, George Lestaris and Ren\'{e} Meusel. The Need for a Versioned Data Analysis Software Environment~\cite{wssspe2_blomer}
\item Ryan Chamberlain and Jennifer Schommer. Using {Docker} to Support Reproducible Research~\cite{wssspe2_chamberlain}
\item Neil Chue Hong. Minimal information for reusable scientific software~\cite{wssspe2_chue_hong}
\item Tom Crick, Benjamin A. Hall and Samin Ishtiaq. ``Can I Implement Your Algorithm?'': A Model for Reproducible Research Software~\cite{wssspe2_crick}
\item Bryan Marker, Don Batory, Field G. Van Zee and Robert van de Geijn. Making Scientific Computing Libraries Forward Compatible~\cite{wssspe2_marker}
\item Stephen Piccolo. Building Portable Analytical Environments to improve sustainability of computational-analysis pipelines in the sciences~\cite{wssspe2_piccolo}
\end{itemize}

\item Code Testing \& Code Review
\begin{itemize}
\item Thomas Clune, Michael Rilee and Damian Rouson. Testing as an Essential Process for Developing and Maintaining Scientific Software~\cite{wssspe2_clune}
\item Marian Petre and Greg Wilson. Code Review For and By Scientists~\cite{wssspe2_petre}
\item Andrew E. Slaughter, Derek R. Gaston, John Peterson, Cody J. Permann, David Andrs and Jason M. Miller. Continuous Integration for Concurrent {MOOSE} Framework and Application Development on {GitHub}~\cite{wssspe2_slaughter}
\end{itemize}

\section{Other parts of the workshop}

WSSSPE2 will include two keynote presentations, and lightning talks from accepted paper authors, in addition to a set of breakout sessions.  The breakouts will allow attendees to participate in discussions about sustainability, future actions, and two of the five areas that resulted from the community submission, review, and grouping process.  There will also be reporting from the breakout groups to collect information for all attendees, as well as people who were not able to attend, to understand the discussions.

@@

@@

>>>>

\todo{update this section with intro to full report - currently just intro for pre-workshop stuff}

link to live notes - \url{https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing}

The Second Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} was held on Sunday, 16 November 2014 in conjunction with the 2014 International Conference for High Performance Computing, Networking, Storage and Analysis (SC14)\footnote{\url{http://sc14.supercomputing.org}}.
WSSSPE2 followed a general initial workshop, WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1}, that was held in 2013 in conjunction with the SC13 conference, and a focused workshop, WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, that was held in 2014 in conjunction with the SciPy conference.

@@

\section{Submissions, Peer-Review, and Peer-Grouping}

\todo{shorten this section? - cite~\cite{WSSSPE2-pre-report} somewhere here}

@@

@@

The organizers then chose the sort that contained five groups as the one that felt most meaningful, then decided on names for the five groups, namely:
\begin{itemize}
\item Exploring Sustainability
\item Software Development Experiences
\item Credit \& Incentives
\item Reproducibility \& Reuse \& Sharing
\item Code Testing \& Code Review
\end{itemize}

Finally, since some of the papers were not represented by cards in the process, they were not placed in groups by the peer-grouping system, so the authors of these papers were asked which group seemed the best for their papers, and those papers were then placed in those groups.  The later sections of this paper that discuss the breakout groups include a list of the papers associated with each group.

\section{Keynotes}

\todo{VOLUNTEER NEEDED to write summaries for Kaitlin and Neil's keynotes}

\section{Defining Sustainability}

\todo{Dan to discuss initial breakout and results}

\section{Exploring Sustainability}

\todo{Colin to lead writing of this section}

Google doc notes - \url{https://docs.google.com/a/uchicago.edu/document/d/10XkshP3YjXnA5JQgP-UFHbOxshWczzehWGSo6V7bNH4/edit}

\todo{short intro to group here}

\subsection{Papers}

\begin{itemize}
\item Mario {Rosado de Souza} and Robert Haines and Caroline Jay. Defining Sustainability through Developers' Eyes: Recommendations from an Interview Study~\cite{wssspe2_rosada_de_souza}
\item Robert Downs, W. Christopher Lenhardt, Erin Robinson, Ethan Davis, Nicholas Weber. Community Recommendations for Sustainable Scientific Software~\cite{wssspe2_downs}
\item Abani Patra, Matthew Jones, Steven Gallo, Kyle Marcus and Tevfik Kosar. Role of Online Platforms, Communications and Workflows in Developing Sustainable Software for Science Communities~\cite{wssspe2_patra}
\item Marlon Pierce, Suresh Marru and Chris Mattmann.{WSSSPE2}: Patching It Up, Pulling It Forward~\cite{wssspe2_pierce}
\item Justin Shi. Seeking the Principles of Sustainable Software Engineering~\cite{wssspe2_shi}
\item Colin C. Venters, Michael K. Griffiths, Violeta Holmes, Rupert R. Ward and David J. Cooke. The Nebuchadnezzar Effect: Dreaming of Sustainable Software through Sustainable Software Architectures~\cite{wssspe2_venters}
\end{itemize}

\subsection{Discussion and Actions}

\section{Software Development Experiences}

\todo{Nancy to lead writing of this section}

Google doc notes - \url{https://docs.google.com/a/uchicago.edu/document/d/1_q0jmiEPNFRGAEtGLk_QxQ1vVokgTDYruTGURowNGkY/edit}

\todo{short intro to group here}

\subsection{Papers}

\begin{itemize}
\item Jordan Adams, Sai Nudurupati, Nicole Gasparini, Daniel Hobley, Eric Hutton, Gregory Tucker and Erkan Istanbulluoglu. Landlab: Sustainable Software Development in Practice ~\cite{wssspe2_adams}
\item Alice Allen and Judy Schmidt. Looking before leaping: Creating a software registry~\cite{wssspe2_allen}
\item Carl Boettiger, Ted Hart, Scott Chamberlain and Karthik Ram. Building software, building community: lessons from the {ROpenSci} project~\cite{wssspe2_boettiger}
\item Michael R. Crusoe and C.Titus Brown. Channeling community contributions to scientific software: a hackathon experience~\cite{wssspe2_crusoe}
\item Yolanda Gil,  Eunyoung Moon and James Howison. No Science Software is an Island: Collaborative Software Development Needs in Geosciences~\cite{wssspe2_gil}
\item Ted Habermann, Andrew Collette, Steve Vincena, Werner Benger, Jay Jay Billings, Matt Gerring, Konrad Hinsen, Pierre de Buyl, Mark K\"{o}nnecke, Filipe Rnc Maia and Suren Byna. The Hierarchical Data Format ({HDF}): A Foundation for Sustainable Data and Software~\cite{wssspe2_habermann}
\item Marcus Hanwell, Patrick O'Leary and Bob O'Bara. Sustainable Software Ecosystems: Software Engineers, Domain Scientists, and Engineers Collaborating for Science~\cite{wssspe2_hanwell}
\item Eric Hutton, Mark Piper, Irina Overeem, Albert Kettner and James Syvitski. Building Sustainable Software - The {CSDMS} Approach~\cite{wssspe2_hutton}
\item W. Christopher Lenhardt, Stanley Ahalt, Matt Jones, J. Aukema, S. Hampton, S. R. Hespanh, R. Idaszak and M. Schildhauer. {ISEES-WSSI} Lessons for Sustainable Science Software from an Early Career Training Institute on Open Science Synthesis~\cite{wssspe2_lenhardt}
\item Jory Schossau and Greg Wilson. Which Sustainable Software Practices Do Scientists Find Most Useful?~\cite{wssspe2_schossau}
\item James S. Spencer, Nicholas S. Blunt, William A. Vigor, Fionn D. Malone, W. M. C. Foulkes, James J. Shepherd and Alex J. W. Thom. The {H}ighly {A}ccurate {N-DE}terminant ({HANDE}) quantum {Monte Carlo} project: Open-source stochastic diagonalisation for quantum chemistry~\cite{wssspe2_spencer}
\end{itemize}

\subsection{Discussion and Actions}

\section{Credit \& Incentives}

\todo{Dan (OR VOLUNTEER) to lead writing of this section}

Google doc notes - \url{https://docs.google.com/document/d/1Jsi_trBT5cjEXVnGjGHJpcIhMEcVpAn6XmYNxfIHXv0/edit#heading=h.gk1dzxu9wi6q}

\todo{short intro to group here}

\subsection{Papers}

\begin{itemize}
\item James Howison. Retract bit-rotten publications: Aligning incentives for sustaining scientific software~\cite{wssspe2_howison}
\item Daniel Katz and Arfon Smith. Implementing Transitive Credit with {JSON-LD}~\cite{wssspe2_katz}
\item Ian Kelley. Publish or perish: the credit deficit to making software and generating data~\cite{wssspe2_kelley}
\end{itemize}

\subsection{Discussion and Actions}

citation ecosystems sub-breakout: \url{https://docs.google.com/document/d/1SOja7k0mXwvOB2tQD3ZW3O38Qv4NwfA21zdMQ9JlUQc/edit}

Taxonomy of Contributorship/Guidelines for Software for Tenure Review sub-breakout: \url{https://docs.google.com/document/d/1sDZLaBR2elZ9Tj6eLIwEtf5VqQ_x_AN_SetKtJlFUpI/edit}

Prizes sub-breakout: \url{https://docs.google.com/document/d/1foSppeHQeQ-qOjd-Rp_CUCNy8JMLxd_vrCDMmqvG2-w/edit} (but as of 2/6, nothing is here and need to look at main breakout doc - Suresh is moving text, I think)

\section{Reproducibility \& Reuse \& Sharing}

\todo{Neil OR VOLUNTEER to lead writing of this section}

google doc notes - \url{https://docs.google.com/document/d/1Mivadj16_9tPrw8Nxmhp72AtNC0BEVO-f7M84FXTbSU/edit#heading=h.elaktnt2tep9}

\todo{short intro to group here}

\subsection{Papers}

\begin{itemize}
\item Jakob Blomer, Dario Berzano, Predrag Buncic, Ioannis Charalampidis, Gerardo Ganis, George Lestaris and Ren\'{e} Meusel. The Need for a Versioned Data Analysis Software Environment~\cite{wssspe2_blomer}
\item Ryan Chamberlain and Jennifer Schommer. Using {Docker} to Support Reproducible Research~\cite{wssspe2_chamberlain}
\item Neil Chue Hong. Minimal information for reusable scientific software~\cite{wssspe2_chue_hong}
\item Tom Crick, Benjamin A. Hall and Samin Ishtiaq. ``Can I Implement Your Algorithm?'': A Model for Reproducible Research Software~\cite{wssspe2_crick}
\item Bryan Marker, Don Batory, Field G. Van Zee and Robert van de Geijn. Making Scientific Computing Libraries Forward Compatible~\cite{wssspe2_marker}
\item Stephen Piccolo. Building Portable Analytical Environments to improve sustainability of computational-analysis pipelines in the sciences~\cite{wssspe2_piccolo}
\end{itemize}

\subsection{Discussion and Actions}

\section{Code Testing \& Code Review}

\todo{Karen to lead writing of this section}

google notes doc - 
\url{https://docs.google.com/document/d/1G1_U5-hkAvDKtTgT4DvUhIYvyYUY4cBBvQHJIdgADa0/edit#heading=h.v7g0nqpn1fur}

\todo{short intro to group here}

\subsection{Papers}

\begin{itemize}
\item Thomas Clune, Michael Rilee and Damian Rouson. Testing as an Essential Process for Developing and Maintaining Scientific Software~\cite{wssspe2_clune}
\item Marian Petre and Greg Wilson. Code Review For and By Scientists~\cite{wssspe2_petre}
\item Andrew E. Slaughter, Derek R. Gaston, John Peterson, Cody J. Permann, David Andrs and Jason M. Miller. Continuous Integration for Concurrent {MOOSE} Framework and Application Development on {GitHub}~\cite{wssspe2_slaughter}
\end{itemize}

\subsection{Discussion and Actions}

@@

\todo{currently just pre-workshop stuff - needs to be updated}

@@

\appendix

\section{Attendees}

\todo{VOLUNTEER NEEDED to neaten this - alphabetize, columnize, add short intro sentence that these were generated from people who signed the google doc saying they were there and are not complete, also added names from breakout groups}

Daniel S. Katz, U Chicago \& Argonne
Gabrielle Allen, U Illinois
Alice Allen, Astrophysics Source Code Library (ASCL)
Neil Chue Hong, Software Sustainability Institute \& University of Edinburgh, N 
Tom Crick, Cardiff Metropolitan University, UK
Samin Ishtiaq, Microsoft Research Cambridge, UK
Matt Turk, NCSA/UIUC
Colin C. Venters, University of Huddersfield
Wolfgang Bangerth
Carl Boettiger, UCSC/ropensci,
Chris Bogart, ISR/CMU
Abani Patra, Univ @ Buffalo
Kyle Marcus, Univ @ Buffalo
Karen Cranston, NESCent
Andre Luckow, Rutgers
Marlon Pierce, Indiana University, 
Frank L\"{o}ffler, Louisiana State University,
Steven R. Brandt, Louisiana State University, 
James Howison, University of Texas
Ray Idaszak, RENCI/UNC,
Chris Lenhardt, 
Robert H. McDonald - Indiana University - Data to Insight Center/Libraries - 
Marcus D. Hanwell, Kitware, 
Tom Clune - NASA GSFC - 
Mike Rilee - NASA/GSFC, Rilee Systems Technologies - 
Sarah Harris, University of Leeds, 
James Spencer, Imperial College London,
John W. Peterson, Idaho National Laboratory,
Cody J Permann, Idaho National Laboratory, 
Richard Littauer, MIT/University of Saarland 
Aleksandra Pawlik, Software Sustainability Institute, 
Jordan Adams, Tulane University, 
Jory Schossau, Michigan State University, 
Robert R Downs, CIESIN, Columbia University, 
Ethan Davis, UCAR Unidata, 
Nancy Wilkins-Diehr, SDSC, 
Suresh Marru, Indiana University 
Hilmar Lapp, National Evolutionary Synthesis Center (NESCent) and Duke University, 
Bryan Marker, The University of Texas at Austin,  
James Shepherd, Rice University, 
Ethan White, University of Florida/Utah State University, 
Mark Schildhauer,  NCEAS, Univ. Calif, Santa Barbara
Carol Song, Purdue University, 
John W. Cobb, 
Matt Jones,  University of California nia Santa Barbara
Erin Robinson, Foundation for Earth Science 
Nick Jones, New Zealand eScience Infrastructure, University of Auckland, 
Yolanda Gil, Information Sciences Institute, University of Southern California, 
Karthik Ram, , University of California, Berkeley. Berkeley Institute for Data Science and rOpenSci
Sushil Prasad, Georgia State University, Atlanta 
Nicole Gasparini, Tulane University, New Orleans
Stephen Piccolo,  Brigham Young University, Utah
David Henty,  EPCC, The University of Edinburgh
Alan Simpson,  EPCC, The University of Edinburgh
Berkin Malkoc, Istanbul Technical University
Pierre-Yves Aquilanti TOTAL E\&P R\&T USA
Greg Wilson  Software Carpentry
Anshu Dubey  Lawrence Berkeley National Laboratory
Sol Greenspan  National Science Foundation
Tracy Teal Data Carpentry
Frank Seinstra Netherlands eScience Center
Dirk Colbry, Michigan State University
Kaitlin Thaney,  Mozilla Science Lab, 
Ted Habermann, The HDF Group
Benjamin Pharr, University of Mississippi, 
Eric Hutton,  University of Colorado
Ray Plante, NCSA/University of Illinois Urbana-Champaign
Kurt Glaesemann Pacific northwest national lab
Ardita Shkurti University of Nottingham
Andy Nutter-Upham, Whitehead Institute,
David Bernholdt, Oak Ridge National Laboratory

Names below are from working groups - probably duplicates of above but perhaps not all...

Colin C. Venters, University of Huddersfield
Erin Robinson, Foundation for Earth Science, 
Robert Downs, CIESIN 
Justin Shi
Ethan Davis 
Kyle Marcus
Andy Nutter-Upham, Whitehead Institute, 
Abani Patra

Chris Lenhardt , Nancy Wilkins-Diehr , Marcus Hanwell , Tracy Teal , Jory Schossau , Andre Luckow

Ian Kelley 
Ray Idaszak 
Ray Plante 
Anshu Dubey
Dirk Colbry 
Dan Katz
Chris Bogart
Sol Greenspan

Kaitlin Thaney
Abani Patri
Frank Seinstra
Frank Loffler
Richard
Robert McDonald
Steven Brandt

Samin Ishtaq, Nathan Weeks, Rupert Nash, Stephen Piccolo, Hilmar Lapp, Tom Crick, Jakob Blomer, Neil Chue Hong

Sarah Harris, Carl Boettiger, Bryan Marker, Berkin Malkoc, Carol Song, Mark Schildhauer, Benjamin Pharr

Karen Cranston, National Evolutionary Synthesis Center,
Greg Wilson, Software Carpentry, 
Wolfgang Bangerth 
Ethan White, Utah State Univ/University of Florida, 
John Peterson, Idaho National Laboratory, jwpeterson@gmail.com, 
David Bernholdt
Aleksandra Pawlik, Software Sustainability Institute, 
Kurt Glaesemann
Cody Permann
Maxime Hughes
Anshu Dubey
Tom Clune
Mike Rilee
